---
title: "Annotation merge"
output: html_notebook
---



```{r}
#input nr data
nr = read.csv("..\\annotation\\meta_nr_unique.tsv", sep = "\t", header = FALSE)
nr = nr[,c(1,2,11,12)]
colnames(nr) = c("transcript_name", "nr_gene_name", "nr_e-value", "nr_bitscore")
```


```{r}
#input swissprot data
sp = read.csv("..\\annotation\\meta_swissprot_unique.tsv", sep = "\t", header = FALSE)
sp = sp[,c(1,2,11,12)]
colnames(sp) = c("transcript_name", "sp_gene_name", "sp_e-value", "sp_bitscore")
```




```{r}
#number of transcripts annotated in nr but not sp
length(setdiff(nr$transcript_name, sp$transcript_name))
```


```{r}
#number of transcripts annotated in sp but not nr
length(setdiff(sp$transcript_name, nr$transcript_name))
```

#Note: this figure (the smaller of the setdiffs) is also equal to the sum of the difference between the number of rows in each dataframe and the larger of the setdiffs. This is because any transcripts not in the setdiffs is double-counted if you were to sum the number of rows in each dataframe, but by subtracting these the double-counted transcripts cancel out. However, the result is smaller than the larger setdiff by an amount equal tot he smaller setdiff because we also subtracted these out, so we can add that number back in to get the larger setdiff.

```{r}
(nrow(nr) - nrow(sp)) + length(setdiff(sp$transcript_name, nr$transcript_name))
```



```{r}
#full outer join dataframes
merged = merge(nr, sp, all = TRUE)
merged
```


```{r}
#we just want one column that has the gene names,start with one preferring the swissprot names
#If the transcript has a swissprot annotation, keep that one. Otherwise, use the NR annotation. Have a column that indicates which reference the annotation is from

pref_sp = merged[c('transcript_name', 'sp_gene_name')]
pref_sp$origin = "swissprot"
colnames(pref_sp) = c("transcript_name", "gene_name", "origin")

#get indices where gene name is currently NA
na = is.na(pref_sp$gene_name)
#set gene name of na entries to the nr gene name
pref_sp$gene_name[na] = merged$nr_gene_name[na]
#set origin of nr gene names to nr
pref_sp$origin[na] = "nr"

#remove the "." and version number on each gene identifier for use with downstream tools
pref_sp$gene_name = sapply(strsplit(pref_sp$gene_name, ".", fixed = TRUE), function(x) x[1])

pref_sp
```


```{r}
#now do the same thing except select the annotation with the higher bitscore (this should be easily adaptable to use the e-value if desired instead but we use the bitscore because it corrects for database size and so can be used to compare hits in databases of different sizes)
bits = merged$nr_bitscore - merged$sp_bitscore
#this problem is a bit more complicated because we will get NAs for any transcript that did not annotate in both references
#For those entries, we can simply select the annotation present in the pref_sp dataframe since there is only one option. Start by copying the dataframe, we will overwrite some entries later
na = is.na(bits)
pref_bit = pref_sp

#now we need to set the annotations of any index with a bitscore difference >0 to the nr annotation
#start by setting any difference greater than 1 to 1, with anything less to NA.
#Now the <0 difference and missing both annotations transcripts are labeled as NA, and anything with a bitscore >0 is labeled as 1
na = ifelse(bits>0, 1, NA)
#convert to boolean
na = is.na(na)
#We now want to overwrite anything that is not NA, in other words anything with a bitscore difference >0 should be set to the nr annotation
#Note: anything with a bitscore difference of 0 (entries with the same bitscore in both databases) will have the swissprot annotation
pref_bit$gene_name[!na] = merged$nr_gene_name[!na]
#now sset origin for those entries
pref_bit$origin[!na] = "nr"

#we are now done because anything that had a bitscore less than one had a better match to the swissprot database, and so it's entry would be the same as the pref_sp dataframe we copied 

pref_bit$bit_diff = bits
pref_bit
```


Search for nr entries here:
https://www.ncbi.nlm.nih.gov/protein/TNN81439.1


```{r}
#write gene_trans_map to file
write.csv(pref_sp, "..\\annotation\\gene_trans_map.csv")
```

```{r}
#input genome swissprot data
sp = read.csv("..\\annotation\\meta_swissprot_unique.tsv", sep = "\t", header = FALSE)
sp = sp[,c(1,2)]
colnames(sp) = c("transcript_name", "gene_name")
sp$gene_name = sapply(strsplit(sp$gene_name, ".", fixed = TRUE), function(x) x[1])
sp$origin = "swissprot"
```

```{r}
#write gene_trans_map to file
write.csv(sp, "..\\annotation\\gene_trans_map.csv")
```


```{r}
#input genome swissprot data
sp = read.csv("..\\annotation\\genome_swissprot.tsv", sep = "\t", header = FALSE)
sp = sp[,c(1,2)]
colnames(sp) = c("transcript_name", "gene_name")
sp$gene_name = sapply(strsplit(sp$gene_name, ".", fixed = TRUE), function(x) x[1])
sp$origin = "swissprot"
```

```{r}
#write genome gene_trans_map to file
write.csv(sp, "..\\annotation\\genome_gene_trans_map.csv")
```






















































































