---
title: "GO analysis"
output: html_notebook
params:
  gtm: ""
  de: ""
  counts: ""
  info: ""
  baseout : ""
  listout : ""
  kegg : 0
---

```{r}
library(ggplot2)
library(tidyr)
library(goseq)
library(lessR)
library(GO.db)
library(plyr)
library(AnnotationDbi)
```



```{r}
#check to make sure output directories exist, create them if not

#match \\ (escaped), then match anything other than a \ (escaped) until the end of the string
list_dir = unlist(strsplit(params$listout, "\\\\[^\\]*$"))
image_dir = unlist(strsplit(params$baseout, "\\\\[^\\]*$"))

if(!dir.exists(list_dir)){
  dir.create(list_dir)
}
if(!dir.exists(image_dir)){
 dir.create(image_dir)
}
```

```{r}
#read in annotation, remove some unneeded columns and then get rid of duplicates
gtm = unique(read.csv(params$gtm)[-c(1,2)])
#select only swissprot entries
gtm = subset(gtm, gtm$origin == "swissprot")
```

```{r}
#read in deg list
degs = read.csv(params$de)
names(degs)[names(degs) == "X"] = "gene_name"
#join gtm with degs to get swissprot degs
degs = merge(degs, gtm)
```


```{r}

assayed_genes = gtm$gene_name
de_genes = degs$gene_name

gene_vector = as.integer(assayed_genes %in% de_genes)
names(gene_vector) = assayed_genes
head(gene_vector)
```


```{r}
#read in count and length data
counts = read.csv(params$counts)
counts$X = sapply(strsplit(counts$X, ".", fixed = TRUE), function(x) x[1])
#set rownames to gene names
rownames(counts) = counts$X
#remove gene names column
counts = counts[,-1]
#We only want the length data, which is the second half of the dataframe
counts = counts[,(length(colnames(counts))/2+1):length(colnames(counts))]
#get vector of named averge length for each gene
ave_len = apply(counts, 1, mean)

#need to keep only swissprot genes
#We do this step now because it is much more efficient once we have relatively small vectors than trying to merge two large dataframes at the start of this chunk
ave_len = ave_len[names(ave_len) %in% gtm$gene_name]
```

```{r}
library(reticulate)
```

```{python}
def extract_go(go_strs):
  temp = [go_str.split("[")[1:] for go_str in go_strs]
  return [[go_str.split("]")[0] for go_str in group] for group in temp]
```



```{r}
#read in annotation info
gene_info = read.csv(params$info)
#only keep id and GO columns
gene_info = gene_info[,c("accession_id", "go.molecular_function.", "go.biological_process.", "go.cellular_component.")]

#process GO columns to only keep GO identifiers. We do this in python because it is much more efficient and significantly easier there because we can use complex list comprehensions
gene_info$go.molecular_function. = py$extract_go(gene_info$go.molecular_function.)
gene_info$go.biological_process. = py$extract_go(gene_info$go.biological_process.)
gene_info$go.cellular_component. = py$extract_go(gene_info$go.cellular_component.)

#create named list in correct format for goseq by combining all go terms for each gene into a single vector
gene2cat = apply(gene_info[,-1], 1, function(x) unlist(x, use.names = FALSE))
names(gene2cat) = gene_info$accession_id

```


#Need to try this with more stringent cutoffs to see if I can get plot to look more normal. Short DEGs are accounting for a lot of the DE proportion

```{r}
#save image
png(filename = paste0(params$baseout, "pwf.png"), width = 650)

pwf = nullp(DEgenes = gene_vector, bias.data = ave_len, plot.fit = TRUE)

dev.off()
```

```{r}
go = goseq(pwf = pwf, gene2cat = gene2cat, test.cats = c("GO:CC", "GO:BP", "GO:MF"), method = "Wallenius")
go_terms = subset(go, go$numDEInCat > 0)
go$over_represented_pvalue = p.adjust(go$over_represented_pvalue, method = "fdr")
go$under_represented_pvalue = p.adjust(go$under_represented_pvalue, method = "fdr")
go = subset(go, go$over_represented_pvalue <= 0.05 | go$under_represented_pvalue <= 0.05)
go
go_terms
```

```{r}
gen_intercept = function(df){
  over = which(df$over_represented_pvalue > 0.05)
  if(length(over) > 0){
    index = nrow(df) - over[1] + 1.5
  }
  else{
    index = 0.5
  }
  return(index)
}
```


```{r}
#KEGG terms
if(params$kegg){
  #read in annotation info
  gene_info = read.csv(params$info)
  #only keep id and KEGG columns
  #Ok to keep as a dataframe because it is a one to one mapping
  gene2cat = gene_info[,c("accession_id", "path_desc")]
  
  go_kegg = goseq(pwf = pwf, gene2cat = gene2cat, method = "Wallenius")
  go_kegg$over_represented_pvalue = p.adjust(go_kegg$over_represented_pvalue, method = "fdr")
  go_kegg$under_represented_pvalue = p.adjust(go_kegg$under_represented_pvalue, method = "fdr")
  go_kegg = subset(go_kegg, go_kegg$over_represented_pvalue <= 0.05 | go_kegg$under_represented_pvalue <= 0.05)
  go_kegg
  
  #if there are signifcantly enriched KEGG pathways
  if(nrow(go_kegg) > 0){
    #save sig list
    write.csv(go_kegg, paste0(params$listout, "sig_KEGG_list.csv"))
    
    go_kegg_pivot = pivot_longer(go_kegg, names_to = 'type', values_to = 'numDE', cols = c('numInCat', 'numDEInCat'))
    go_kegg_pivot$type = factor(go_kegg_pivot$type, levels = c("numInCat", "numDEInCat"))
    go_kegg_pivot$category = factor(go_kegg_pivot$category, levels = go_kegg$category)

    ggplot(data = go_kegg_pivot, aes(x = forcats::fct_rev(category), y = numDE)) + geom_bar(aes(fill = type), position = "dodge", stat = "identity") + ggtitle("KEGG pathway composition") + theme(axis.text.x = element_text(angle = 0, hjust=1)) + ylab("Number of genes") + scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) + coord_flip() + xlab("KEGG category") + geom_vline(xintercept = gen_intercept(go_kegg))

  ggsave(paste0(params$baseout, "sig_KEGG.png"), width = 8, height = 11, units = "in")
  }
}
```



```{r}
#GO composition DE list
MF_all_terms = subset(go_terms, go_terms$ontology == "MF", select = c(term, numDEInCat, category))
BP_all_terms = subset(go_terms, go_terms$ontology == "BP", select = c(term, numDEInCat, category))
CC_all_terms = subset(go_terms, go_terms$ontology == "CC", select = c(term, numDEInCat, category))
sum(MF_all_terms$numDEInCat)
```


```{r}
splitCount = function(x) { x[1]/length(x)}
```


```{r}
getParents = function(parents, child_terms){
  #map GO terms to parent terms
  parents = as.list(parents)
  temp = parents[child_terms$category]
  #get rid of relationship name
  temp[] = lapply(temp, unname)
  #convert to dataframe
  temp = stack(temp)
  colnames(temp) = c("parent", "child")
  #combine parent terms with count information (number of DE genes in that category)
  temp = merge(temp, child_terms, by.x = "child", by.y = "category")
  #if a child has multiple parents, split the DE count between parents (so for a child with 2 parents and 1 DE gene, each parent should have 0.5 DE gene counts from it). This will keep the column total the same and prevent some categories from becoming inflated
  num_split = ddply(temp, .(child), numcolwise(splitCount))
  #replace old numDEIncat column with new one
  temp = merge(subset(temp, select = -c(numDEInCat)), num_split)
  #since there are duplicate parents, group by parents and calculate the total number of DE genes for each parent term. This will discard all non-numeric columns other than the parent column (term, child), but we don't need them anymore anyway
  temp = ddply(temp, .(parent), numcolwise(sum))
  #get description information for the parent terms
  temp$definition = AnnotationDbi::select(GO.db, keys = unique(temp$parent), columns = "TERM", keytype = "GOID")$TERM
  colnames(temp) = c("category", "numDEInCat", "term")
  return(temp)
}
```

```{r}
numAncestors = function(ancestors, curr){
  #map GO terms to ancestors
  ancestors = as.list(ancestors)
  temp = ancestors[curr$category]
  #get rid of relationship name
  temp[] = lapply(temp, unname)
  num_A = stack(lapply(temp, length))
  colnames(num_A) = c("num", "category")
  #decrement num column by 1 to account for "all" term at end
  num_A$num = num_A$num-1
  return(num_A)
}
```

```{r}
numParents = function(parents, curr){
  #map GO terms to parents
  parents = as.list(parents)
  temp = parents[curr$category]
  #get rid of relationship name
  temp[] = lapply(temp, unname)
  num_P = stack(lapply(temp, length))
  colnames(num_P) = c("num", "category")
  return(num_P)
}
```


#If a GO term is two-level deep but has a large number of parents, it will appear to be deep than it acutally is because they are all listed as ancestors and will be stepped up. Need to subtract the number of parents from the number of ancestors to get "real" depth
```{r}
even_GO_Depth = function(lineage, terms_df, depth = 2){
  if(lineage == "MF"){
    parents = GOMFPARENTS
    ancestors = GOMFANCESTOR
  }
  else if(lineage == "BP"){
    parents = GOBPPARENTS
    ancestors = GOBPANCESTOR
  }
  else{
    parents = GOCCPARENTS
    ancestors = GOCCANCESTOR
  }
  #get the number of ancestors each term has
  num_A = numAncestors(ancestors, terms_df)
  #get the number of parents each term has
  num_P = numParents(parents, terms_df)
  #adjust number of ancestors to account for multiple parents. If there is more than one parent, subtract that number - 1 from the number of ancestors since the parents are all at the same level and would otherwise make the go term appear deeper in the graph than it truly is. While this solution will not detect this problem until the ancestor GO terms are in the immediate parent level, it is more robust than not having it and things that are split too much are likely to end up grouped in the other category or washed out by stronger signals anyway
  num_A$num = num_A$num - (num_P$num - 1)
  #subset df to keep those rows with less than the desired number of ancestors ancestors
  keep = terms_df[which(num_A$num <= depth),]
  #subset df into those GO terms that need to be decreased to the desired number of ancestors (higher terms on the graph)
  step_up = terms_df[which(num_A$num > depth),]
  while(nrow(step_up) > 0){
    #Bring GO terms in step_up up one level
    step_up = getParents(parents, step_up)
    #combine new rows with the ones left from before
    temp = rbind(keep, step_up)
    #group categories together and combine the number of DE genes in them
    temp = ddply(temp, .(category), numcolwise(sum))
    #get description information for the terms
    temp$definition = AnnotationDbi::select(GO.db, keys = unique(temp$category), columns = "TERM", keytype = "GOID")$TERM
    colnames(temp) = c("category", "numDEInCat", "term")
    #calculate number of ancestors for new terms
    num_A = numAncestors(ancestors, temp)
    keep = temp[which(num_A$num <= depth),]
    step_up = temp[which(num_A$num > depth),]
  }
  #combine categories with <2% of the total number of genes into an "Other" category
  keep = temp[which(temp$numDEInCat >= 0.02*sum(temp$numDEInCat)),]
  oth = temp[which(temp$numDEInCat < 0.02*sum(temp$numDEInCat)),]
  temp = rbind(keep, data.frame(category = "Other", numDEInCat = sum(oth$numDEInCat), term = "Other"))
  return(temp)
}
```

This is currently set to only output significant enrichment pie charts since the others did not turn out to be very useful

```{r}
#only run this if there are GO terms in the dataframe
if(nrow(MF_all_terms) > 0){
  #PieChart(x = term, y = numDEInCat, hole = 0, values = "%", data = MF_all_terms, main = "Molecular Function GO composition of total DE list", pdf_file = paste0(params$baseout, "all_MF.pdf"), width = 12.5)
  
  #MF_all_even = even_GO_Depth("MF", MF_all_terms, depth = 2)
#  if(nrow(MF_all_even) > 0){
    #PieChart(x = term, y = numDEInCat, hole = 0, values = "%", data = MF_all_even, main = "Molecular Function GO composition of total DE list at 2-term depth", pdf_file = paste0(params$baseout, "all_even_MF.pdf"), width = 12.5)
#  }
}
```

```{r}
#GO composition significantly enriched DE list
MF_sig_terms = subset(go, go$ontology == "MF")
BP_sig_terms = subset(go, go$ontology == "BP")
CC_sig_terms = subset(go, go$ontology == "CC")
```


```{r}
if(nrow(MF_sig_terms) > 0){
  MF_sig_pivot = pivot_longer(MF_sig_terms, names_to = 'type', values_to = 'numDE', cols = c('numInCat', 'numDEInCat'))
  MF_sig_pivot$type = factor(MF_sig_pivot$type, levels = c("numInCat", "numDEInCat"))
  MF_sig_pivot$term = factor(MF_sig_pivot$term, levels = MF_sig_terms$term)
  ggplot(data = MF_sig_pivot, aes(x = forcats::fct_rev(term), y = numDE)) + geom_bar(aes(fill = type), position = "dodge", stat = "identity") + ggtitle("Molecular Function GO composition") + theme(axis.text.x = element_text(angle = 0, hjust=1)) + ylab("Number of genes") + scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) + coord_flip() + xlab("GO term") + geom_vline(xintercept = gen_intercept(MF_sig_terms))


  ggsave(paste0(params$baseout, "sig_MF.png"), width = 8, height = 11, units = "in")
  
  #write out sig list
  write.csv(MF_sig_terms, paste0(params$listout, "MF_sig_GO_list.csv"))
}
```

```{r}
if(nrow(BP_all_terms) > 0){
  #PieChart(x = term, y = numDEInCat, hole = 0, values = "%", data = BP_all_terms, main = "Biological Process GO composition of total DE list", pdf_file = paste0(params$baseout, "all_BP.pdf"), width = 12.5)
  
  #BP_all_even = even_GO_Depth("BP", BP_all_terms, depth = 2)

 # if(nrow(BP_all_even) > 0){
    #PieChart(x = term, y = numDEInCat, hole = 0, values = "%", data = BP_all_even, main = "Biological Process GO composition of total DE list at 2-term depth", pdf_file = paste0(params$baseout, "all_even_BP.pdf"), width = 12.5)
#  }
}
```



```{r}
if(nrow(BP_sig_terms) > 0){
  BP_sig_pivot = pivot_longer(BP_sig_terms, names_to = 'type', values_to = 'numDE', cols = c('numInCat', 'numDEInCat'))
  BP_sig_pivot$type = factor(BP_sig_pivot$type, levels = c("numInCat", "numDEInCat"))
  BP_sig_pivot$term = factor(BP_sig_pivot$term, levels = BP_sig_terms$term)
  ggplot(data = BP_sig_pivot, aes(x = forcats::fct_rev(term), y = numDE)) + geom_bar(aes(fill = type), position = "dodge", stat = "identity") + ggtitle("Biological Process GO composition") + theme(axis.text.x = element_text(angle = 0, hjust=1)) + ylab("Number of genes") + scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) + coord_flip() + xlab("GO term") + geom_vline(xintercept = gen_intercept(BP_sig_terms))

  ggsave(paste0(params$baseout, "sig_BP.png"), width = 8, height = 11, units = "in")
  
  write.csv(BP_sig_terms, paste0(params$listout, "BP_sig_GO_list.csv"))
}
```

```{r}
if(nrow(CC_all_terms) > 0){
  #PieChart(x = term, y = numDEInCat, hole = 0, values = "%", data = CC_all_terms, main = "Cellular Component GO composition of total DE list", pdf_file = paste0(params$baseout, "all_CC.pdf"), width = 12.5)
  
  #CC_all_even = even_GO_Depth("CC", CC_all_terms, depth = 3)

 # if(nrow(CC_all_even) > 0){
    #PieChart(x = term, y = numDEInCat, hole = 0, values = "%", data = CC_all_even, main = "Cellular Component GO composition of total DE list at 3-term depth", pdf_file = paste0(params$baseout, "all_even_CC.pdf"), width = 12.5)
#  }
}
```


```{r}
if(nrow(CC_sig_terms) > 0){
  CC_sig_pivot = pivot_longer(CC_sig_terms, names_to = 'type', values_to = 'numDE', cols = c('numInCat', 'numDEInCat'))
  CC_sig_pivot$type = factor(CC_sig_pivot$type, levels = c("numInCat", "numDEInCat"))
  CC_sig_pivot$term = factor(CC_sig_pivot$term, levels = CC_sig_terms$term)
  ggplot(data = CC_sig_pivot, aes(x = forcats::fct_rev(term), y = numDE)) + geom_bar(aes(fill = type), position = "dodge", stat = "identity") + ggtitle("Cellular Component GO composition") + theme(axis.text.x = element_text(angle = 0, hjust=1)) + ylab("Number of genes") + scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) + coord_flip() + xlab("GO term") + geom_vline(xintercept = gen_intercept(CC_sig_terms))


  ggsave(paste0(params$baseout, "sig_CC.png"), width = 8, height = 11, units = "in")
  
  write.csv(CC_sig_terms, paste0(params$listout, "CC_sig_GO_list.csv"))
}
```











